Perturbing SU($n$)-symmetric interactions

### 2020-02-02 collective_sun/sun_perturbation.tex
added code to check multibody eigenvalue problem
rearranged some math

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_VM_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q} \sp{\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{kp} w_{Okq} w_{Okq} + h_{kq} w_{Opk}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv 2 h_{pq} - h_p - h_q,
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv \sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{kp} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

We first simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_k \sum_{p<q} h_{pq} v_k \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where without loss of generality we consider a single-body
perturbation $\V_1$ built out of only one single-spin operator $X$.
The sum in \eqref{eq:H_V1_psi_start} has terms with $k\in\set{p,q}$,
and terms with $k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$,
the permutation operator $\Pi_{pq}$ commutes with $X_k$ and
annihilates on $\ket\psi$, leaving terms at each fixed $k$ of the form
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \sum_{p<q} h_{pq} - \sum_{\substack{p<q\\k\in\set{p,q}}} h_{pq}
  = E_0 - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$; and
$h_k \equiv \sum_\ell h_{k\ell}$.  The terms in
\eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_k h_k v_k X_k \ket\psi
\end{align}
while the terms in \eqref{eq:H_V1_psi_start} with $k\in\set{p,q}$ take
the form
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_k X_q \ket\psi
  + \sum_{p<k} h_{pk} v_k X_p \ket\psi
  = \sum_{k,q} h_{kq} v_k X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_p \p{\sum_{q} h_{pq} v_q - h_p v_p}
  X_p \ket\psi.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_M \ket\psi
  = \sum_{\substack{\p{p,q}\in\C_N\p{2}\\k\in\C_N\p{M}}} h_{pq} w_k
  \Pi_{pq} O_k \ket\psi,
  \label{eq:H_VM_psi_start}
\end{align}
where without loss of generality we consider an $M$-body perturbation
$\V_M$ built out of only one $M$-spin operator $O$.  The sum in
\eqref{eq:H_VM_psi_start} has terms with $p,q\in k$, terms with
$p,q\notin k$, and terms with only one of $p$ or $q\in k$.  The
permutation operator $\Pi_{pq}$ acts trivially on $O_k\ket\psi$ when
$p,q\in k$ or $p,q\notin k$, leaving terms at each fixed $k$ of the
form
\begin{align}
  \sum_{\substack{\p{p,q}\in\C_N\p{2}\\p,q\in k~\t{or}~p,q\notin k}} h_{pq}
  = \sum_{\p{p,q}\in\C_N\p{2}} h_{pq}
  - \sum_{p\in k} \sum_{\substack{q\in\ZZ_N\\q\notin k}} h_{pq}
  = E_0 - \sum_{p\in k}
  \p{\sum_{q\in\ZZ_N} h_{pq} - \sum_{q\in k} h_{pq}},
\end{align}
where we first split a sum over $\p{p,q}\in\C_N\p{2}$ with the
restriction that $p,q\in k~\t{or}~p,q\notin k$ into an unrestricted
sum minus a remainder, and then similarly split the sum over all spins
$q$ with the restriction that $q\notin k$ into a sum over all $q$
minus the remainder $q\in k$.  We thus find that
\begin{align}
  \sum_{\substack{\p{p,q}\in\C_N\p{2}\\p,q\in k~\t{or}~p,q\notin k}} h_{pq}
  = E_0 + \bar h_k,
  &&
  \bar h_k \equiv 2\sum_{\p{p,q}\subset k} h_{pq} - \sum_{p\in k} h_p,
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
where the sum over $\p{p,q}\subset k$ denotes a sum over distinct
choices of two elements in $k$.  The terms in
\eqref{eq:H_VM_psi_start} with only one of $p\in k$ or $q\in k$,
meanwhile, are
\begin{align}
  \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{q\in k}
  h_{pq} w_k \Pi_{pq} O_k \ket\psi
  = \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \sum_{j=1}^M h_{p k_j} w_k \Pi_{p k_j} O_k \ket\psi.
\end{align}
Defining
$k_{jp}\equiv\p{k_1,k_2,\cdots,k_{j-1},p,k_{j+1},\cdots,k_M}$, i.e.~an
element of $\C_N\p{M}$ in which the $j$-th number $k_j$ has been
replaced by $p$, we can simplify
\begin{align}
  \Pi_{p k_j} O_k \ket\psi = O_{k_{jp}}\ket\psi,
\end{align}
and switch the labels for $p$ and $k_j$ to find that
\begin{align}
  \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{q\in k}
  h_{pq} w_k  \Pi_{pq} O_k \ket\psi
  = \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \sum_{j=1}^M h_{pk_j} w_{k_{jp}} O_k \ket\psi.
\end{align}
Defining
\begin{align}
  \p{h\circ w}_k
  \equiv \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{j=1}^M
  h_{pk_j} w_{k_{jp}}
  = \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \p{h_{pk_1} w_{pk_2k_3\cdots k_M}
    + h_{pk_2} w_{k_1pk_3\cdots k_M}
    + \cdots + h_{pk_M} w_{k_1k_2\cdots k_{M-1} p}},
\end{align}
we thus find that
\begin{align}
  H_0 \V_M \ket\psi
  = E_0 \V_M \ket\psi + \sum_{k\in\C_N\p{M}}
  \sp{\bar h_k w_k + \p{h\circ w}_k} O_k \ket\psi.
\end{align}
If $M=1$, then $\bar h_k=-h_k$ and
$\p{h\circ w}_k=\sum_\ell h_{k\ell}w_\ell$, so
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi + \sum_{k\in\ZZ_N}
  \p{\sum_\ell h_{k\ell} w_\ell - h_k w_k} O_k \ket\psi.
\end{align}
If $M=2$, then
$\p{h\circ w}_{pq}=\sum_k\p{h_{kp} w_{kq}+h_{kq}w_{pk}}$, so
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi + \sum_{p<q}
  \sp{\bar h_{pq} w_{pq} + \sum_k\p{h_{kp} w_{kq} + h_{kq} w_{pk}}}
  O_k \ket\psi,
\end{align}
where $\bar h_{pq}=2h_{pq}-h_p-h_q$ with $h_p\equiv\sum_qh_{pq}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analytical results in translationally invariant systems}
\label{sec:trans_inv}

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \bar h_{pq} w_{pq}
  + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  &&
  \bar h_{pq} \equiv 2h_{pq} - h_p - h_q,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
which can be written in matrix form as
\begin{align}
  \bar{\m h} * \m w + \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h}
  = \Delta \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$, $\bar{\m h}$, and $\m w$ are matrices of all $h_{pq}$,
$\bar h_{pq}$ and $w_{pq}$; $\bar{\m h}*\m w$ denotes an element-wise
(Kronecker) product of $\bar{\m h}$ and $\m w$, i.e.~with matrix
elements $\bar h_{pq}w_{pq}$; and
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices.  In order to find solutions to \eqref{eq:cond_2_ref} when
the interaction Hamiltonian $H_0$ is translationally invariant, in
which case $h_{pq}$ and $\bar h_{pq}$ depend only on the separation
$\pm\p{p-q}$, we expand all matrices in a basis that the respects
translational symmetry.  To this end, for all lattice displacements
$d\in\ZZ_L^D$ and wavenumbers $k\in\ZZ_L^D\times2\pi/L$ we define the
{\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms
of these cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
and expand
\begin{align}
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
  &&
  \m h = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \bar{\m h} = \sum_d \bar h_{d,0} \m m_{d,0}.
\end{align}
We can then write the condition in \eqref{eq:cond_2_ref} as a new set
of conditions
\begin{align}
  \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{d\ne0} \bar h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k = \Delta \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a fully symmetric
state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a
fixed-excitation-energy perturbation $\V_2^\Delta$ is then performed
by projecting each $k$-component of $\v{\m w}$ onto the corresponding
$\Delta$-eigenspace of $\ut{\check{\m h}}_k$.  If the perturbation
$\V_2$ is translationally invariant, then the $k$-components of
$\v{\m w}$ for all $k\ne 0$ all vanish, so we only need to consider
the case of $k=0$.

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_VM_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{kp} w_{Okq} w_{Okq} + h_{kq} w_{Opk}}
  - \bar h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{kp} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}
    - \bar h_{pq} \bra{\set{p,q}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

We first simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_k \sum_{p<q} h_{pq} v_k \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where without loss of generality we consider a single-body
perturbation $\V_1$ built out of only one single-spin operator $X$.
The sum in \eqref{eq:H_V1_psi_start} has terms with $k\in\set{p,q}$,
and terms with $k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$,
the permutation operator $\Pi_{pq}$ commutes with $X_k$ and
annihilates on $\ket\psi$, leaving terms at each fixed $k$ of the form
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \sum_{p<q} h_{pq} - \sum_{\substack{p<q\\k\in\set{p,q}}} h_{pq}
  = E_0 - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$; and
$h_k \equiv \sum_\ell h_{k\ell}$.  The terms in
\eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_k h_k v_k X_k \ket\psi
\end{align}
while the terms in \eqref{eq:H_V1_psi_start} with $k\in\set{p,q}$ take
the form
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_k X_q \ket\psi
  + \sum_{p<k} h_{pk} v_k X_p \ket\psi
  = \sum_{k,q} h_{kq} v_k X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_p \p{\sum_{q} h_{pq} v_q - h_p v_p}
  X_p \ket\psi.
  \label{eq:H_V1_psi_end}
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_M \ket\psi
  = \sum_{\substack{\p{p,q}\in\C_N\p{2}\\k\in\C_N\p{M}}} h_{pq} w_k
  \Pi_{pq} O_k \ket\psi,
  \label{eq:H_VM_psi_start}
\end{align}
where without loss of generality we consider an $M$-body perturbation
$\V_M$ built out of only one $M$-spin operator $O$.  The sum in
\eqref{eq:H_VM_psi_start} has terms with $p,q\in k$, terms with
$p,q\notin k$, and terms with only one of $p$ or $q\in k$.  The
permutation operator $\Pi_{pq}$ acts trivially on $O_k\ket\psi$ when
$p,q\in k$ or $p,q\notin k$, leaving terms at each fixed $k$ of the
form
\begin{align}
  \sum_{\substack{\p{p,q}\in\C_N\p{2}\\p,q\in k~\t{or}~p,q\notin k}} h_{pq}
  = \sum_{\p{p,q}\in\C_N\p{2}} h_{pq}
  - \sum_{p\in k} \sum_{\substack{q\in\ZZ_N\\q\notin k}} h_{pq}
  = E_0 - \sum_{p\in k}
  \p{\sum_{q\in\ZZ_N} h_{pq} - \sum_{q\in k} h_{pq}},
\end{align}
where we first split a sum over $\p{p,q}\in\C_N\p{2}$ with the
restriction that $p,q\in k~\t{or}~p,q\notin k$ into an unrestricted
sum minus a remainder, and then similarly split the sum over all spins
$q$ with the restriction that $q\notin k$ into a sum over all $q$
minus the remainder $q\in k$.  We thus find that
\begin{align}
  \sum_{\substack{\p{p,q}\in\C_N\p{2}\\p,q\in k~\t{or}~p,q\notin k}} h_{pq}
  = E_0 - \bar h_k,
  &&
  \bar h_k \equiv \sum_{p\in k}\p{h_p - \sum_{q\in k}h_{pq}},
  &&
  h_p \equiv \sum_q h_{pq}.
\end{align}
The terms in \eqref{eq:H_VM_psi_start} with only one of $p\in k$ or
$q\in k$, meanwhile, are
\begin{align}
  \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{q\in k}
  h_{pq} w_k \Pi_{pq} O_k \ket\psi
  = \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \sum_{j=1}^M h_{p k_j} w_k \Pi_{p k_j} O_k \ket\psi.
\end{align}
Defining
$k_{jp}\equiv\p{k_1,k_2,\cdots,k_{j-1},p,k_{j+1},\cdots,k_M}$, i.e.~an
element of $\C_N\p{M}$ in which the $j$-th number $k_j$ has been
replaced by $p$, we can simplify
\begin{align}
  \Pi_{p k_j} O_k \ket\psi = O_{k_{jp}}\ket\psi,
\end{align}
and switch the labels for $p$ and $k_j$ to find that
\begin{align}
  \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{q\in k}
  h_{pq} w_k  \Pi_{pq} O_k \ket\psi
  = \sum_{k\in\C_N\p{M}} \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \sum_{j=1}^M h_{pk_j} w_{k_{jp}} O_k \ket\psi.
\end{align}
Defining
\begin{align}
  \p{h\circ w}_k
  \equiv \sum_{\substack{p\in\ZZ_N\\p\notin k}} \sum_{j=1}^M
  h_{pk_j} w_{k_{jp}}
  = \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  \p{h_{k_1 p} w_{pk_2k_3\cdots k_M}
    + h_{k_2 p} w_{k_1pk_3\cdots k_M}
    + \cdots + h_{k_M p} w_{k_1k_2\cdots k_{M-1} p}},
\end{align}
we thus find that
\begin{align}
  H_0 \V_M \ket\psi
  = E_0 \V_M \ket\psi + \sum_{k\in\C_N\p{M}}
  \sp{\p{h\circ w}_k - \bar h_k w_k} O_k \ket\psi.
\end{align}
If $M=1$, then $\p{h\circ w}_k=\sum_\ell h_{k\ell}w_\ell$ and
$\bar h_k=h_k$, so
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi + \sum_{k\in\ZZ_N}
  \p{\sum_{\ell\in\ZZ_N} h_{k\ell} w_\ell - h_k w_k} O_k \ket\psi,
\end{align}
which is precisely what we found in \eqref{eq:H_V1_psi_end}.  If
$M=2$, then $\p{h\circ w}_{pq}=\sum_k\p{h_{kp} w_{kq}+h_{kq}w_{pk}}$,
so
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi + \sum_{\p{p,q}\in\C_N\p{2}}
  \sp{\sum_{k\in\ZZ_N}\p{h_{kp} w_{kq} + h_{kq} w_{pk}}
    - \bar h_{pq} w_{pq}}
  O_k \ket\psi,
\end{align}
where $\bar h_{pq}=h_p+h_q-2h_{pq}$ with $h_p\equiv\sum_qh_{pq}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making use of translational invariance}
\label{sec:trans_inv}

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  - \bar h_{pq} w_{pq}
  = \Delta w_{pq},
  &&
  \bar h_{pq} \equiv h_p + h_q - 2h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
which can be written in matrix form as
\begin{align}
  \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h} - \bar{\m h} * \m w
  = \Delta \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$, $\bar{\m h}$, and $\m w$ are matrices of all $h_{pq}$,
$\bar h_{pq}$ and $w_{pq}$;
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices; and $\bar{\m h}*\m w$ denotes an element-wise (Kronecker)
product of $\bar{\m h}$ and $\m w$, i.e.~with matrix elements
$\bar h_{pq}w_{pq}$.  In order to find solutions to
\eqref{eq:cond_2_ref} when the interaction Hamiltonian $H_0$ is
translationally invariant, in which case $h_{pq}$ and $\bar h_{pq}$
depend only on the separation $\pm\p{p-q}$, we expand all matrices in
a basis that the respects translational symmetry.  To this end, for
all lattice displacements $d\in\ZZ_L^D$ and wavenumbers
$k\in\ZZ_L^D\times2\pi/L$ we define the {\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms
of these cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
and expand
\begin{align}
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
  &&
  \m h = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \bar{\m h} = \sum_d \bar h_{d,0} \m m_{d,0}.
\end{align}
We can then write the condition in \eqref{eq:cond_2_ref} as a new set
of conditions
\begin{align}
  \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  - \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  - \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c}
  - \sum_{d\ne0} \bar h_{d,0} \op{d},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k = \Delta \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a fully symmetric
state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a
fixed-excitation-energy perturbation $\V_2^\Delta$ is then performed
by projecting each $k$-component of $\v{\m w}$ onto the corresponding
$\Delta$-eigenspace of $\ut{\check{\m h}}_k$.  If the perturbation
$\V_2$ is translationally invariant, then the $k$-components of
$\v{\m w}$ for all $k\ne 0$ all vanish, so we only need to consider
the case of $k=0$.
